Model Recognition as Planning

### 2018-06-21 doc/aaai-19/model-recognition.tex
goals added to the models recognition formalism

\maketitle
\begin{abstract} 
Given a partially observed plan execution and a set of possible planning models, {\em model recognition} is the task of identifying which model in the set produced the observed execution. The paper formalizes the {\em model recognition} task and proposes a novel method to assess the probability of a given \strips\ model to produce a partially observed plan execution. This method, that we called {\em model recognition as planning}, is robust to missing data in the intermediate states and actions of the observed plan execution besides, it is computable with an off-the-shelf classical planner. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different {\em Turing Machines}. We show that {\em model recognition as planning} succeeds to identify the executed {\em Turing Machine} despite the actual applied transitions, internal machine state or the values of several tape cells, are unknown.
\end{abstract}

@@

Diverse approaches has been proposed for plan recognition such as {\em rule-based systems}, {\em parsing}, {\em graph-covering}, {\em Bayesian nets}, etc~\cite{carberry2001techniques}. {\em Plan recognition as planning} is the model-based approach for plan recognition~\cite{ramirez2012plan,ramirez2009plan}. This approach assumes that the action model of the observed agent is known and leverages it to compute the most likely goal according to the observed plan execution.

@@

\section{Background}
\label{sec:background}
This section formalizes classical planning and characterizes the space of \strips\ models.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\subseteq\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that, when executed starting from the initial state $I$, induces the {\em state trajectory} $\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and $a_i$ ({\small $1\leq i\leq n$}) is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$. A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$; i.e.~if the goal condition is satisfied in the last state resulting from the execution of the plan $\pi$ in the initial state $I$. A solution plan is {\em optimal} if it has minimum length.

\subsection{The observation model}
The observation of a plan $\pi$ executed for solving a given classical planning problem $P=\tup{F,A,I,G}$ is an interleaved combination $\tau=\tup{s_1, a_1, \ldots , a_n, s_m}$ of a {\em partially-observed} sequence of actions and a sequence of {\em partially-observed} states. Because of the partial observability $0\leq n\leq |\pi|$ and $0\leq m\leq |\pi|+1$ and hence, the transitions between two consecutive observed states in $\tau$ may involve the execution of more than a single action. Formally $\theta(s_i,\tup{a_1,\ldots,a_k})=s_{i+1}$, where $k\geq 1$ is unknown and unbound. This means that having $\tau$ does not implies knowing the actual length of $\pi$.

We say that a given observation $\tau$ is {\em compliant} with a given plan $\pi$ iff:
\begin{itemize}
\item The sequence of actions $\tup{a_1, \ldots, a_n}$ in $\tau$ is the same sequence of actions $\pi$ but with certain actions omitted.
\item The sequence of states $\tup{s_1, \ldots, s_m}$ in $\tau$ is the same sequence of states traversed by $\pi$ but (1), with certain states omitted and/or (2), the value of certain fluents omitted in that states, i.e.~$|s_i|\leq |F|$ for every $0\leq i\leq m$.
\end{itemize}

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\tau}$ where:
\begin{itemize}
\item $P=\tup{F,A,I,G}$ is a classical planning problem such that the dynamics of the actions in $A$ is unknown (i.e. the $App(s)$ and $\theta(s,a)$ functions are undefined).
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite and non-empty set of action models, each defining a different dynamics (different $App(s)$ and $\theta(s,a)$ functions for the actions in $A$).
\item $\tau$ is the observation of the execution of a classical plan for solving $P$.
\end{itemize}
A solution to the {\em model recognition} task is a discrete probability distribution $P(\mathcal{M}|\tau)$ that expresses for each model $\mathcal{M}\in M$ its probability of producing $\tau$.

According to the {\em Bayes} rule, the probability of an hypothesis $\mathcal{H}$, provided the observation $\mathcal{O}$, is given by the expression $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In our scenario, hypotheses are about the possible action models models $\mathcal{M}\in M$ while the given observation is the partially observed plan execution $\tau$. The $P(\mathcal{M}|\tau)$ probability distribution can then be estimated in three steps:
\begin{enumerate}
\item Estimating the {\em a priori} probabilities. $P(\tau)$, that measures how surprising is the given observation and $P(\mathcal{M})$, that expresses if one model is a priori more likely than the others. 
\item Estimating the conditional probability $P(\tau|\mathcal{M})$.  Our approach is to estimate this value according to the minimum amount of {\em edition} required by $\mathcal{M}$ to produce a plan $\pi^*_\tau$ such that:
\begin{enumerate}
\item $\pi^*_\tau$ is an optimal solution for the classical planning problem $P$. 
\item $\pi^*_\tau$ is compliant with the given observation $\tau$.
\end{enumerate}
\item Applying the Bayes rule to obtain the normalized posterior probabilities. Since $P(\mathcal{M}|\tau)$ is a probability distribution these probabilities must sum 1 for all the $\mathcal{M}\in M$.
\end{enumerate}

We assume that the observed agent is acting rationally, like in {\em plan recognition as planning}~\cite{ramirez2012plan,ramirez2009plan}. A related approach is recently followed in {\em model reconciliation}~\cite{Kambhampati:mreconciliation:ijcai17} where model edition is used to conform the different PDDL models of two different agents.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we study the particular instantiation of the model recognition task where the dynamics of the actions $A\in P$ are specified with \strips\ action schemas. Figure~\ref{fig:stack} shows the PDDL code for the {\em stack} actions, taken from a four-operator {\em blocksworld}~\cite{slaney2001blocks}.

@@

The size of the space of possible \strips\ models for an action schema $\xi$ is given by the expression, $2^{2|F_{\xi}|}$. Note that \strips\ constraints require negative effects appearing as preconditions and that they cannot be positive effects and also, that positive effects cannot appear as preconditions. For the mentioned {\em blocksworld}, $2^{2|F_{\tt stack}|}=4194304$ while $2^{2|F_{\tt pickup}|}=1024$. We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff both share the same space of possible \strips\ models. For instance, we claim that blocksworld schemas {\small\tt stack} and {\small\tt unstack} are {\em comparable} while  {\small\tt stack} and {\small\tt pickup} are not. Last but not least, two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

@@

\begin{definition}
  Given $\tau$, an observation of the execution of a classical plan for solving $P$ and $\mathcal{M}$, a \strips\ action model built from $F_v$. The {\bf observation edit distance}, $\delta(\mathcal{M},\tau)$, is the minimal edit distance from $\mathcal{M}$ to any {\em comparable} model $\mathcal{M}'$ s.t. $\mathcal{M}'$ produces a plan $\pi^*_\tau$ that is optimal for $P$ and is compliant with $\tau$; \[\delta(\mathcal{M},\tau)=\min_{\forall \mathcal{M}' \rightarrow \tau} \delta(\mathcal{M},\mathcal{M}')\]
\end{definition}

Our $\delta(\mathcal{M},\tau)$ distance could also be defined assessing the edition required by the observed plan execution to match the given model. This implies defining {\em edit operations} that modify $\tau$ instead of $\mathcal{M}$~\cite{sohrabi:precognition:IJCAI2016}. Our definition of the {\em observation edit distance} is more practical since normally $F_v$ is smaller than $F$. In practice, the number of {\em variable objects} is normally smaller than the number of objects in the observed states.

\subsection{The \strips\ $P(\mathcal{M}|\tau)$ probability distribution}
Since we assume that the dynamics of the actions $A\in P$ are specified with \strips\ action schemas, we can compute the $P(\mathcal{M}|\tau)$ probability as follows:
\begin{enumerate}
\item If we assume that a priori all models are equiprobable because there are no reasons to assume that one is a priori more likely than the others then $P(\mathcal{M})=\frac{1}{\prod_{\xi\in\mathcal{M}} 2^{2|F_v(\xi)|}}$.
\item With respect to $P(\tau)$, if we assume that all the observations of plan executions with a maximum of $n$ observed actions and $m$ observed states are equiprobable then $P(\tau)=\frac{1}{2^{n\times|A|}\times2^{m\times|F|}}$.
\item To estimate the conditional probability $P(\tau|\mathcal{M})$ we compute the {\em observation edit distance} $\delta(\mathcal{M},\tau)$ and map it into a $[0,1]$ likelihood with the expression, $1-\frac{\delta(\mathcal{M},\tau)}{\delta(\mathcal{M},*)}$.
\end{enumerate}

\section{Model recognition as planning}
This section shows that, when the dynamics of the actions $A\in P$ are specified with \strips\ action schemas, then the {\em observation edit distance} can be computed with a compilation of a classical planning with conditional effects. The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$ in the observation $\tau$}. The solution plan continues with a postfix that validates the edited model $\mathcal{M}'$ on the given observations $\tau$.
\end{enumerate}

Figure~\ref{fig:plan-pdistance} shows the plan for editing a given {\em blockswold} model where the positive effects {\tt\small (handempty)} and {\tt\small (clear ?v1)} of the {\tt\small stack} schema are missing. The edited action model is validated at the observation of a four action plan for inverting a two-block tower where: states $s_0$ and $s_4$ are fully observed while the intermediate states, $s_1$, $s_2$ and $s_3$, are unobserved. 
\begin{figure}
{\tt\scriptsize
00 : (insert\_add\_handempty\_stack)\\
01 : (insert\_add\_clear\_stack\_var1)\\
02 : (apply\_unstack blockB blockA i1 i2)\\
03 : (apply\_putdown blockB i2 i3)\\
04 : (apply\_pickup blockA i3 i4)\\
05 : (apply\_stack blockA blockB i4 i5)\\
06 : (validate\_1)
}
 \caption{\small Plan for editing (steps [0-1]) and validating (steps [2-6]) a given \strips\ planning model for the {\em blockswold}.}
\label{fig:plan-pdistance}
\end{figure}

Note that our interest is not in $\mathcal{M}'$, the edited model resulting from the compilation, but in the number of required {\em edit operations} (insertions and deleitions) required by $\mathcal{M}'$ to be validated in the given observation, e.g. $\delta(\mathcal{M},\tau)=2$ for the example in Figure~\ref{fig:plan-pdistance}. In this case $\delta(\mathcal{M},*)=3\times 2\times (11+5)$ since there are 4 action schemes ({\small\tt pickup}, {\small\tt putdown}, {\small\tt stack} and {\small\tt unstack}) and $|F_v|=|F_v(stack)|=|F_v(unstack)|=11$ while $|F_v(pickup)|=|F_v(putdown)|=5$. %The {\em observation edit distance} is exactly computed if the classical planning task resulting from our compilation is optimally solved (according to the number of edit actions); is approximated if it is solved with a satisfying planner; and is a less accurate estimate (but faster to be computed) if the solved task is a relaxation of the classical planning task that results from our compilation~\cite{bonet2001planning}.

@@

\subsection{The compilation formalization}
Given a \strips\ model $\mathcal{M}\in M$ and the observation of a plan execution $\tau$, our compilation outputs a classical planning task $P'=\tup{F',A',I',G'}$ such that:
\begin{itemize}
\item $F'$ contains:
\begin{itemize}
\item The original fluents $F$. 
\item Fluents modeling the space of \strips\ models. That is, for every $f\in F_v(\xi)$, the fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$ that represent the edited action model.
\item The fluents $F_{\pi}=\{plan(name(a_i),\Omega^{ar(a_i)},i)\}_{\small 1\leq i\leq n}$ to code the $i^{th}$ action in $\tau$. The static facts $next_{i,i+1}$ and the fluents $at_i$, {\small $1\leq i< n$}, are also added to iterate through the $n$ steps of $\tau$.
\item The fluents $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\tau$ where the action model is validated.
\item The fluents $mode_{edit}$ and $mode_{val}$ to indicate whether the operator schemas are edited or validated.
\end{itemize}
\item $I'$ extends the original initial state $I$ with the fluent $mode_{edit}$ set to true as well as the fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i<n$}, for tracking the plan step where the action model is validated. Our compilation assumes that initially $\mathcal{M}'$ is defined as $\mathcal{M}$. Therefore fluents $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ hold as given by $\mathcal{M}$.

\item $G'=G\bigcup\{at_n,test_m\}$.
\item $A'$ comprises three kinds of actions with conditional effects:
\begin{enumerate}
\item Actions for {\em editing} operator schema $\xi\in\mathcal{M}$:
\begin{itemize}
\item Actions for {\bf removing} a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programPre_{f,\xi}})=&\{\neg del_{f}(\xi),\neg add_{f}(\xi), mode_{edit}, pre_{f}(\xi)\},\\
\cond(\mathsf{programPre_{f,\xi}})=&\{\emptyset\}\rhd\{\neg pre_{f}(\xi)\}.
\end{align*}
\end{small}

\item Actions for {\bf adding} a {\em negative} or {\em positive} effect $f\in F_v(\xi)$ to the action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programEff_{f,\xi}})=&\{\neg del_{f}(\xi),\neg add_{f}(\xi), mode_{edit}\},\\
\cond(\mathsf{programEff_{f,\xi}})=&\{pre_{f}(\xi)\}\rhd\{del_{f}(\xi)\},\\
&\{\neg pre_{f}(\xi)\}\rhd\{add_{f}(\xi)\}.
\end{align*}
\end{small}
\end{itemize}
Besides these actions, $A$ also contains the actions for {\em inserting} a precondition and for {\em deleting} a negative/positive effect.

@@

A classical planning frame $\Phi=\tup{F,A}$ can encode the {\em transition function} $\delta$ of a {\em Turing Machine} $\mathcal{M}$ as follows:
\begin{itemize}
\item Fluents $F$ are instantiated from a set of four {\em predicates} $\Psi$: {\small\tt (head ?x)} that encodes the current position of the header in the tape. {\small\tt (next ?x1 ?x2)} encoding that the cell {\tt ?x2} follows cell {\tt ?x1} in the tape. {\small\tt (symbol-$\sigma$ ?x)} encoding that the tape cell {\tt ?x} contains the symbol $\sigma\in\Sigma$. {\small\tt (state-$q$)} encoding that $q\in Q$ is the current machine state. Given a set of {\em objects} $\Omega$ that represent the cells in the tape of the given Turing Machine, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of the predicates in $\Psi$.
\item Actions $A$ are instantiated from \strips\ operator schema. For each transition in $\delta$, a \strips\ schema is defined:
\begin{itemize}
\item The {\bf header} is {\small\tt transition-id(?xl ?x ?xr)} where $id$ uniquely identifies the transition in $\delta$. Parameters $?xl$, $?x$ and $?xr$ are tape cells.
\item The {\bf preconditions} are {\small\tt(head ?x)} and {\small\tt (next ?xl ?x) (next ?x ?xr)} to make $?x$ the tape cell pointed by the header and $?xl/?xr$ its left/right neighbours. Preconditions {\small\tt(symbol-$\sigma$ ?x)} and {\small\tt (state-$q$)} are also included to capture the symbol currently pointed by the header and the current machine state.
\item The {\bf delete effects} remove the symbol currently pointed by the header and the currrent machine state while the {\bf positive effects} set the new symbol pointed by the header and the new machine state according to $\delta$.
\end{itemize}
\end{itemize}

@@

Since the {\em transition function} of a {\em Turing Machine} can be encoded as a classical planning frame $\Phi=\tup{F,A}$, executions of that {\em Turing Machine} are definable as {\em plan traces} $\tau=\tup{s_0,a_1,s_1,\ldots,a_n,s_n}$.

>>>>

\maketitle
\begin{abstract} 
Given a partially observed plan execution and a set of possible planning models, {\em model recognition} is the task of identifying the model in the set that produced the observed plan execution. The paper formalizes the {\em model recognition} task and proposes a novel method to assess the probability of a \strips\ model to produce a given partially observed plan execution. This method, that we called {\em model recognition as planning}, is robust to missing data in the intermediate states and actions of the observed plan execution besides, it is computable with an off-the-shelf classical planner. The effectiveness of {\em model recognition as planning} is shown in a set of \strips\ models encoding different {\em Turing Machines}. We show that {\em model recognition as planning} succeeds to identify the executed {\em Turing Machine} despite the actual applied transitions, internal machine state or the values of several tape cells, are unknown.
\end{abstract}

@@

Diverse approaches has been proposed for plan recognition such as {\em rule-based systems}, {\em parsing}, {\em graph-covering}, {\em Bayesian nets}, etc~\cite{carberry2001techniques}. {\em Plan recognition as planning} is the model-based approach for plan recognition~\cite{ramirez2012plan,ramirez2009plan}. This approach assumes that the action model of the observed agent is known and leverages it to compute the most likely goal of the agent according to the observed plan execution.

@@

\section{Background}
\label{sec:background}
This section formalizes classical planning and the observation of the execution of a classical plan.

@@

A {\em classical planning problem} is a tuple $P=\tup{F,A,I,G}$, where $I$ is an initial state and $G\subseteq\mathcal{L}(F)$ is a goal condition. A {\em plan} for $P$ is an action sequence $\pi=\tup{a_1, \ldots, a_n}$ that, when executed starting from the initial state $I$, induces the {\em state trajectory} $\tup{s_0, s_1, \ldots, s_n}$ such that $s_0=I$ and $a_i$ ({\small $1\leq i\leq n$}) is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The {\em plan length} is denoted with $|\pi|=n$.

A plan $\pi$ {\em solves} $P$ iff $G\subseteq s_n$; i.e.~if the goal condition is satisfied in the last state resulting from the execution of the plan $\pi$ in the initial state $I$. A solution plan is {\em optimal} if it has minimum length.

\subsection{The observation model}
Given classical planning problem $P=\tup{F,A,I,G}$, the observation of the execution of a plan $\pi$ that solves $P$ is an interleaved combination $\tau=\tup{s_1, a_1, \ldots , a_n, s_m}$ of a {\em partially-observed} sequence of actions and a sequence of {\em partially-observed} states. Because of the partial observability $0\leq n\leq |\pi|$ and $0\leq m\leq |\pi|+1$ and hence, the transitions between two consecutive observed states in $\tau$ may involve the execution of more than a single action. Formally $\theta(s_i,\tup{a_1,\ldots,a_k})=s_{i+1}$, where $k\geq 1$ is unknown and unbound. This means that having $\tau$ does not implies knowing the actual length of $\pi$.

We say that a given observation $\tau$ is {\em compliant} with a given plan $\pi$ iff:
\begin{itemize}
\item The sequence of actions $\tup{a_1, \ldots, a_n}$ in $\tau$ is a subsequence of $\pi$.
\item The sequence of states $\tup{s_1, \ldots, s_m}$ in $\tau$ is the same sequence of states traversed by $\pi$ but (1), with certain states omitted and/or (2), the value of certain fluents omitted in that states, i.e.~$|s_i|\leq |F|$ for every $0\leq i\leq m$.
\end{itemize}

\section{Model Recognition}
\label{sec:recognition}
The {\em model recognition} task is a tuple $\tup{P,M,\tau}$ where:
\begin{itemize}
\item $P=\tup{F,A,I,G}$ is a classical planning problem such that the dynamics of the actions in $A$ is unknown because the $App(s)$ and $\theta(s,a)$ functions are undefined.
\item $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_m\}$ is a finite and non-empty set of action models, each defining a different dynamics (a different $App(s)$ and $\theta(s,a)$ pair) for the actions in $A$.
\item $\tau$ is the observation of the execution of a solution plan $\pi$ for $P$.
\end{itemize}
A solution to the {\em model recognition} task is the discrete probability distribution $P(\mathcal{M}|\tau)$ that expresses, for each model $\mathcal{M}\in M$, its probability of producing the observation $\tau$.

According to the {\em Bayes} rule, the probability of an hypothesis $\mathcal{H}$, provided the observation $\mathcal{O}$, is given by the expression $P(\mathcal{H}|\mathcal{O})=\frac{P(\mathcal{O}|\mathcal{H})P(\mathcal{H})}{P(\mathcal{O})}$. In the {\em model recognition} task, hypotheses are about the possible action models $\mathcal{M}\in M$ while the given observation is the partially observed plan execution $\tau$. The $P(\mathcal{M}|\tau)$ probability distribution can then be estimated in three steps:
\begin{enumerate}
\item Computing the {\em a priori} probabilities. $P(\tau)$, that measures how surprising is the given observation and $P(\mathcal{M})$, that expresses if one model is a priori more likely than the others. 
\item Computing the conditional probability $P(\tau|\mathcal{M})$.  Our approach is to estimate this value according to the minimum amount of {\em edition} that is required by $\mathcal{M}$ to produce a plan $\pi^*_\tau$ such that:
\begin{enumerate}
\item $\pi^*_\tau$ is an optimal solution for the classical planning problem $P$. 
\item $\pi^*_\tau$ is compliant with the given observation $\tau$.
\end{enumerate}
\item Applying the Bayes rule to obtain the normalized posterior probabilities (the $P(\mathcal{M}|\tau)$ probabilities must sum 1 for all the $\mathcal{M}\in M$).
\end{enumerate}

Note that the second step assumes that the observed agent is acting rationally, like in {\em plan recognition as planning}~\cite{ramirez2012plan,ramirez2009plan}. A related approach is recently followed for {\em model reconciliation}~\cite{Kambhampati:mreconciliation:ijcai17} where model edition is used to conform the different PDDL models of two different agents.

\section{Recognition of \strips\ models}
\label{sec:asPlanning}
Here we analyze the particular instantiation of the {\em model recognition} task where the dynamics of the actions $A\in P$ (i.e. the $App(s)$ and $\theta(s,a)$ functions) are specified with \strips\ action schemas. Figure~\ref{fig:stack} shows the PDDL code for the {\em stack} actions, taken from a four-operator {\em blocksworld}~\cite{slaney2001blocks}.

@@

The size of the space of possible \strips\ models for an action schema $\xi$ is given by the expression, $2^{2|F_{\xi}|}$ (\strips\ constraints require negative effects appearing as preconditions and that they cannot be positive effects and also, that positive effects cannot appear as preconditions). For the mentioned {\em blocksworld}, $2^{2|F_{\tt stack}|}=4194304$ while $2^{2|F_{\tt pickup}|}=1024$. We say that two \strips\ schemes $\xi$ and $\xi'$ are {\em comparable} iff both share the same space of possible \strips\ models. For instance, we claim that blocksworld schemas {\small\tt stack} and {\small\tt unstack} are {\em comparable} while  {\small\tt stack} and {\small\tt pickup} are not. Last but not least, two \strips\ models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff there exists a bijective function $\mathcal{M} \mapsto \mathcal{M}^*$ that maps every action schema $\xi\in\mathcal{M}$ to a comparable schema $\xi'\in\mathcal{M'}$ and vice versa.

@@

\begin{definition}
  Given $\tau$, an observation of the execution of a plan for solving $P$ and $\mathcal{M}$, a \strips\ action model built from $F_v$. The {\bf observation edit distance}, $\delta(\mathcal{M},\tau)$, is the minimal edit distance from $\mathcal{M}$ to any {\em comparable} model $\mathcal{M}'$ s.t. $\mathcal{M}'$ produces a plan $\pi^*_\tau$ optimal for $P$ and compliant with $\tau$; \[\delta(\mathcal{M},\tau)=\min_{\forall \mathcal{M}' \rightarrow \tau} \delta(\mathcal{M},\mathcal{M}')\]
\end{definition}

Our $\delta(\mathcal{M},\tau)$ distance could also be defined assessing the edition required by the observed plan execution to match the given model. This implies defining {\em edit operations} that modify $\tau$ instead of $\mathcal{M}$~\cite{sohrabi:precognition:IJCAI2016}. Our definition of the {\em observation edit distance} is more practical since normally $F_v$ is smaller than $F$. In practice, the number of {\em variable objects} is normally smaller than the number of objects in a planning problem.

\subsection{The $P(\mathcal{M}|\tau)$ probability for \strips\ models}
Since we assume that the dynamics of the actions $A\in P$ is specified with \strips\ action schemas, we can estimate the $P(\mathcal{M}|\tau)$ probability as follows:
\begin{enumerate}
\item Assuming that a priori, all action models $\mathcal{M}\in M$ are equiprobable because there are no reasons to assume that one is a priori more likely than the others. Therefore, $P(\mathcal{M})=\frac{1}{\prod_{\xi\in\mathcal{M}} 2^{2|F_{\xi}|}}$.
\item Assuming that all the observations of plan executions with a maximum of $n$ observed actions and $m$ observed states are equiprobable then $P(\tau)=\frac{1}{2^{n\times|A|}\times2^{m\times|F|}}$.
\item To estimate the conditional probability $P(\tau|\mathcal{M})$ we compute the {\em observation edit distance} $\delta(\mathcal{M},\tau)$ and map it into a $[0,1]$ likelihood with the expression, $1-\frac{\delta(\mathcal{M},\tau)}{\delta(\mathcal{M},*)}$.
\end{enumerate}

\section{Model recognition as planning}
This section shows that, when the dynamics of the actions $A\in P$ are specified with \strips\ action schemas, then $\delta(\mathcal{M},\tau)$ can be computed with a compilation of a classical planning with conditional effects. The intuition behind this compilation is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item {\bf Edits the action model $\mathcal{M}$ to build $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that modifies the preconditions and effects of the action schemes in $\mathcal{M}$ using to the two {\em edit operations} defined above, {\em deletion} and {\em insertion}. 
\item {\bf Validates the edited model $\mathcal{M}'$ in the observation $\tau$}. The solution plan continues with a postfix that validates the edited model $\mathcal{M}'$ on the given observations $\tau$.
\end{enumerate}

Figure~\ref{fig:plan-pdistance} shows the plan for editing a given {\em blockswold} model where the positive effects {\tt\small (handempty)} and {\tt\small (clear ?v1)} of the {\tt\small stack} schema are missing. The edited action model is validated at the observation of a four action plan for inverting a two-block tower where the intermediate states, $s_1$, $s_2$ and $s_3$, are unobserved. 
\begin{figure}
{\tt\scriptsize
00 : (insert\_add\_handempty\_stack)\\
01 : (insert\_add\_clear\_stack\_var1)\\
02 : (apply\_unstack blockB blockA i1 i2)\\
03 : (apply\_putdown blockB i2 i3)\\
04 : (apply\_pickup blockA i3 i4)\\
05 : (apply\_stack blockA blockB i4 i5)\\
06 : (validate\_1)
}
 \caption{\small Plan for editing (steps [0-1]) and validating (steps [2-6]) a given \strips\ planning model for the {\em blockswold}.}
\label{fig:plan-pdistance}
\end{figure}

Note that our interest is not in $\mathcal{M}'$, the edited model resulting from the compilation, but in the number of required {\em edit operations} (insertions and deletions) required by $\mathcal{M}'$ to be validated in the given observation, e.g. $\delta(\mathcal{M},\tau)=2$ for the example in Figure~\ref{fig:plan-pdistance}. In this case $\delta(\mathcal{M},*)=3\times 2\times (11+5)$ since there are 4 action schemes ({\small\tt pickup}, {\small\tt putdown}, {\small\tt stack} and {\small\tt unstack}) and $|F_v|=|F_{\tt stack}|=|F_{\tt unstack}|=11$ while $|F_{\tt pickup}|=|F_{\tt putdown}|=5$. %The {\em observation edit distance} is exactly computed if the classical planning task resulting from our compilation is optimally solved (according to the number of edit actions); is approximated if it is solved with a satisfying planner; and is a less accurate estimate (but faster to be computed) if the solved task is a relaxation of the classical planning task that results from our compilation~\cite{bonet2001planning}.

@@

\subsection{The compilation formalization}
Given a \strips\ model $\mathcal{M}\in M$ and the observation $\tau$ of the execution of a plan for solving $P=\tup{F,A,I,G}$, our compilation outputs a classical planning task $P'=\tup{F',A',I',G'}$ such that:
\begin{itemize}
\item $F'$ contains:
\begin{itemize}
\item The original fluents $F$. 
\item Fluents $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ modeling the space of \strips\ models. 
\item The fluents $F_{\pi}=\{plan(name(a_i),\Omega^{ar(a_i)},i)\}_{\small 1\leq i\leq n}$ to code the $i^{th}$ action in $\tau$. The static facts $next_{i,i+1}$ and the fluents $at_i$, {\small $1\leq i< n$}, are also added to iterate through the $n$ steps of $\tau$.
\item The fluents $\{test_j\}_{1\leq j\leq m}$, indicating the state observation $s_j\in\tau$ where the action model is validated.
\item The fluents $mode_{edit}$ and $mode_{val}$ to indicate whether the operator schemas are edited or validated.
\end{itemize}
\item $I'$ extends the original initial state $I$ with the fluent $mode_{edit}$ set to true as well as the fluents $F_{\pi}$ plus fluents $at_1$ and $\{next_{i,i+1}\}$, {\small $1\leq i<n$}, for tracking the plan step where the action model is validated. Our compilation assumes that initially $\mathcal{M}'$ is defined as $\mathcal{M}$. Therefore fluents $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ hold as given by $\mathcal{M}$.

\item $G'=G\bigcup\{at_n,test_m\}$.
\item $A'$ comprises three kinds of actions with conditional effects:
\begin{enumerate}
\item Actions for {\em editing} operator schema $\xi\in\mathcal{M}$:
\begin{itemize}
\item Actions for adding a {\em precondition} $f\in F_v(\xi)$ from the action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programPre_{f,\xi}})=&\{\neg pre_{f}(\xi), \neg del_{f}(\xi),\neg add_{f}(\xi), mode_{edit}\},\\
\cond(\mathsf{programPre_{f,\xi}})=&\{\emptyset\}\rhd\{pre_{f}(\xi)\}.
\end{align*}
\end{small}

\item Actions for adding a {\em negative} or {\em positive} effect $f\in F_v(\xi)$ to the action schema $\xi\in\mathcal{M}$.

\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{programEff_{f,\xi}})=&\{\neg del_{f}(\xi),\neg add_{f}(\xi), mode_{edit}\},\\
\cond(\mathsf{programEff_{f,\xi}})=&\{pre_{f}(\xi)\}\rhd\{del_{f}(\xi)\},\\
&\{\neg pre_{f}(\xi)\}\rhd\{add_{f}(\xi)\}.
\end{align*}
\end{small}
\end{itemize}
Besides these actions, the $A'$ set also contains the actions for {\em deleting} a precondition and a negative/positive effect.

@@

A classical planning frame $\Phi=\tup{F,A}$ can encode the {\em transition function} $\delta$ of a {\em Turing Machine} $\mathcal{M}$ as follows:
\begin{itemize}
\item Fluents $F$ are instantiated from a set of four {\em predicates} $\Psi$: {\small\tt (head ?x)} that encodes the current position of the header in the tape. {\small\tt (next ?x1 ?x2)} encoding that the cell {\tt ?x2} follows cell {\tt ?x1} in the tape. {\small\tt (symbol-$\sigma$ ?x)} encoding that the tape cell {\tt ?x} contains the symbol $\sigma\in\Sigma$. {\small\tt (state-$q$)} encoding that $q\in Q$ is the current machine state. Given a set of {\em objects} $\Omega$ that represent the cells in the tape of the given Turing Machine, the set of fluents $F$ is induced by assigning objects in $\Omega$ to the arguments of the predicates in $\Psi$.
\item Actions $A$ are instantiated from \strips\ operator schema. For each transition in $\delta$, a \strips\ schema is defined:
\begin{itemize}
\item The {\bf header} is {\small\tt transition-id(?xl ?x ?xr)} where $id$ uniquely identifies the transition in $\delta$. Parameters $?xl$, $?x$ and $?xr$ are tape cells.
\item The {\bf preconditions} are {\small\tt(head ?x)} and {\small\tt (next ?xl ?x) (next ?x ?xr)} to make $?x$ the tape cell pointed by the header and $?xl/?xr$ its left/right neighbors. Preconditions {\small\tt(symbol-$\sigma$ ?x)} and {\small\tt (state-$q$)} are also included to capture the symbol currently pointed by the header and the current machine state.
\item The {\bf delete effects} remove the symbol currently pointed by the header and the current machine state while the {\bf positive effects} set the new symbol pointed by the header and the new machine state according to $\delta$.
\end{itemize}
\end{itemize}

@@

Since the {\em transition function} of a {\em Turing Machine} can be encoded as a classical planning frame $\Phi=\tup{F,A}$, executions of that {\em Turing Machine} are definable as an action sequence $\tup{a_1, \ldots, a_n}$ that induces the {\em state trajectory} $\tup{s_0, s_1, \ldots, s_n}$ such that $a_i$ ({\small $1\leq i\leq n$}) is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$.

\subsection{Experimental setup}
We randomly generated a $M=\{\mathcal{M}_1,\ldots,\mathcal{M}_{100}\}$ set of one-hundred different {\em Turing Machines} where each $\mathcal{M}\in M$ is a seven-symbol six-state {\em Turing Machine}. The classical planning frame $\Phi=\tup{F,A}$ is the same for all the {\em Turing Machines}, there is an $a\in A$ action for each pair of tape symbol and non-terminal state machine, while the $\theta(s,a)$ function is defined differently for each the machines (using a different \strips\ action model).
